\appendix

\chapter{Common Traps (Master List)}

\section{Chapter 0: What AI Actually Is}

\begin{itemize}
    \item Treating AI outputs as truth
    \item Assuming AI understands context
    \item ``It works on my test set, ship it!''
    \item Anthropomorphizing the model
\end{itemize}

\section{Chapter 1: Python \& Data}

\begin{itemize}
    \item Not looking at your data
    \item Trusting data providers
    \item Ignoring missing data patterns
    \item Not versioning data
\end{itemize}

\section{Chapter 2: Math You Can't Escape}

\begin{itemize}
    \item Memorizing formulas without understanding
    \item Getting stuck in math rabbit holes
    \item Skipping linear algebra
    \item Treating probability as just counting
\end{itemize}

\section{Chapter 3: Classical ML}

\begin{itemize}
    \item Not using cross-validation
    \item Tuning hyperparameters on the test set
    \item Ignoring class imbalance
    \item Forgetting about feature scaling
\end{itemize}

\section{Chapter 4: Neural Networks}

\begin{itemize}
    \item Not normalizing inputs
    \item Using sigmoid for hidden layers
    \item Not shuffling data
    \item Forgetting to set model to eval mode
    \item Not checking for NaNs
\end{itemize}

\section{Chapter 5: Transformers \& LLMs}

\begin{itemize}
    \item Trusting LLM outputs without verification
    \item Using LLMs for tasks requiring reasoning
    \item Ignoring cost
    \item Not handling edge cases
\end{itemize}

\section{Chapter 6: Modern AI Systems}

\begin{itemize}
    \item Over-relying on LLMs
    \item Not versioning prompts
    \item Ignoring latency
    \item No fallback logic
\end{itemize}

\section{Chapter 7: Production AI}

\begin{itemize}
    \item Deploying and forgetting
    \item Optimizing for accuracy alone
    \item Not planning for retraining
    \item Adding AI because it's trendy
\end{itemize}

\chapter*{Final Thoughts}
\addcontentsline{toc}{chapter}{Final Thoughts}

You've now seen AI from first principles: not as magic, but as optimization, pattern matching, and engineering tradeoffs.

\textbf{Remember}:
\begin{itemize}
    \item AI is powerful but narrow
    \item Data quality matters more than algorithm choice
    \item Models are tools, not solutions
    \item Production is 90\% unglamorous infrastructure
    \item Sometimes the best AI is no AI
\end{itemize}

\textbf{Next steps}:
\begin{enumerate}
    \item Build the mini projects. Experience beats reading.
    \item Read papers, but focus on intuition over proofs.
    \item Deploy something small to production. Feel the pain.
    \item Join communities (forums, Discord, conferences). Learn from practitioners.
    \item Stay skeptical. Question hype. Demand evidence.
\end{enumerate}

Good luck. The field needs developers who understand AI deeply—not just how to call APIs, but how to build, debug, and deploy robust intelligent systems.

Now go build something real.

\vspace{1cm}

\noindent\textit{This guide is in the spirit of OSTEP: pragmatic, skeptical, and focused on understanding over hype. For feedback or questions, open an issue on GitHub.}
